Mortar-based entropy stable discontinuous Galerkin methods

### 2018-10-19 docs/ESDG_mortar/main.tex

\paragraph{Hybrid meshes:} It was shown in \cite{chan2018skew} that an entropy stable

@@

We can now define operators which map between surface and mortar quadrature points.  Let $\bm{E}_f$ denote the map from surface to mortar points, and let $\tilde{\bm{E}}_f $ denote the map from mortar to surface points.  Both operators are defined through an $L^2$ projection to the trace space and interpolation to appropriate points
\[
\bm{E}_f = \bm{T}_m \bm{P}_f, \qquad \tilde{\bm{E}}_f = \bm{T}_f \bm{P}_m.
\]  
We also define a mortar boundary matrix
\[
\tilde{\bm{B}}_i = \bm{W}_m\diag{\hat{\bm{n}}_i}.
\]
Then, a mortar-based formulation can be given as follows:
\begin{equation}
\bm{M}\td{\bm{u}_N}{t} + \begin{bmatrix} \bm{V}_q \\ \bm{V}_f \\ \bm{V}_m \end{bmatrix}^T
\LRp{\begin{bmatrix}
\bm{Q}_i-\bm{Q}_i^T & \bm{E}^T\bm{B}_i &\\
-\bm{B}_i\bm{E} &  & \bm{B}_i\tilde{\bm{E}}_f \\
& -\tilde{\bm{B}}_i{\bm{E}}_f & 
\end{bmatrix} \circ \bm{F}_S}\bm{1} + \bm{V}_m^T\tilde{\bm{B}}_i\bm{f}_i^* = 0.  
\label{eq:mortar}
\end{equation}
Here, we have appended an extra row and column to the decoupled SBP matrix, and the inter-element numerical flux $\bm{f}^*$ is now computed in terms of the \emph{mortar} nodes.

We first note that
\begin{align*}
\bm{B}_i\tilde{\bm{E}}_f &= \diag{\bm{W}_f} \bm{T}_f \bm{P}_m = \diag{\hat{\bm{n}}} \bm{W}_f \bm{T}_f \bm{M}_m^{-1}\bm{T}_m\bm{W}_m\\
&=  \bm{W}_f \bm{T}_f \bm{M}_m^{-1}\bm{T}^T_m\bm{W}_m\diag{\hat{\bm{n}}_i} = \bm{P}_f^T\bm{T}^T_m \tilde{\bm{B}}_i = \LRp{\tilde{\bm{B}}_i{\bm{E}}_f}^T
\end{align*}
where we have used that $\bm{W}_m$ is diagonal and that the outward normal $\hat{\bm{n}}_i$ on the reference element is constant over each face.  This implies skew-symmetry of the decoupled SBP operator and entropy stability of the overall system.

@@

While the formulation presented above is convenient for analysis, it is computationally expensive to implement.  However, using properties of the mortar matrices $\bm{E}_f, \tilde{\bm{E}}_f$, we can rewrite the above formulation in a way which reflects a traditional mortar-based finite element implementation.  In other words, we wish to implement (\ref{eq:mortar}) such that the only modification from the implementations in \cite{chan2017discretely} is a pre-processing step on mortar faces.

We first note that, since $\bm{E}_f$ maps from surface nodes to mortar nodes, $\bm{V}_m = \bm{E}_f \bm{V}_f$.  Thus, we can rewrite the numerical flux contribution as
\begin{align*}
 \bm{V}_m^T\tilde{\bm{B}}_i\bm{f}_i^* &= \bm{V}_f^T \bm{E}_f^T \bm{W}_m\diag{\hat{\bm{n}}_i} \bm{f}_i^* = \bm{V}_f^T  \bm{P}_f^T\bm{T}_m^T\bm{W}_m\diag{\hat{\bm{n}}_i} \bm{f}_i^*\\
 &= \bm{V}_f^T \bm{W}_f \diag{\hat{\bm{n}}_i} \bm{T}_m\bm{M}_m^{-1}\bm{T}_m^T\bm{W}_m \bm{f}_i^* =  \bm{V}_f^T \bm{B}_i \tilde{\bm{E}}_f\bm{f}_i^*.
\end{align*}

We can similarly reformulate the remaining mortar contributions within (\ref{eq:mortar}).  Multiplying out

%\bm{M}\td{\bm{u}_N}{t} + \begin{bmatrix} \bm{V}_q \\ \bm{V}_f \\ \bm{V}_m \end{bmatrix}^T
%\LRp{\begin{bmatrix}
%\bm{Q}_i-\bm{Q}_i^T & \bm{E}^T\bm{B}_i &\\
%-\bm{B}_i\bm{E} &  & \bm{B}_i\tilde{\bm{E}}_f \\
%& -\tilde{\bm{B}}_i{\bm{E}}_f & 
%\end{bmatrix} \circ \bm{F}_S}\bm{1} + \bm{V}_m^T\tilde{\bm{B}}_i\bm{f}_i^* = 0.

@@

>>>>

\paragraph{Hybrid meshes:} entropy stable DG methods on hybrid meshes were introduced in \cite{chan2018skew} using a skew-symmetric formulation.  The resulting methods are stable for more arbitrary choices of surface quadrature, in particular when an SBP property may not hold.

@@

We can now define operators which map between surface and mortar quadrature points.  Let $\bm{E}_m$ denote the map from surface to mortar points, and let $\tilde{\bm{E}}_m $ denote the map from mortar to surface points.  Both operators are defined through an $L^2$ projection to the trace space and interpolation to appropriate points
\[
\bm{E}_m = \bm{T}_m \bm{P}_f, \qquad \tilde{\bm{E}}_m = \bm{T}_f \bm{P}_m.
\]  
We also define a mortar boundary matrix
\[
\tilde{\bm{B}}_i = \bm{W}_m\diag{\hat{\bm{n}}_i}.
\]
Then, a mortar-based formulation can be given as follows:
\begin{equation}
\bm{M}\td{\bm{u}_N}{t} + \begin{bmatrix} \bm{V}_q \\ \bm{V}_f \\ \bm{V}_m \end{bmatrix}^T
\LRp{\begin{bmatrix}
\bm{Q}_i-\bm{Q}_i^T & \bm{E}^T\bm{B}_i &\\
-\bm{B}_i\bm{E} &  & \bm{B}_i\tilde{\bm{E}}_m \\
& -\tilde{\bm{B}}_i{\bm{E}}_m & 
\end{bmatrix} \circ \bm{F}_S}\bm{1} + \bm{V}_m^T\tilde{\bm{B}}_i\bm{f}_i^* = 0.  
\label{eq:mortar}
\end{equation}
Here, we have appended an extra row and column to the decoupled SBP matrix, and the inter-element numerical flux $\bm{f}^*$ is now computed in terms of the \emph{mortar} nodes.

We first note that
\begin{align*}
\bm{B}_i\tilde{\bm{E}}_m &= \diag{\bm{W}_f} \bm{T}_f \bm{P}_m = \diag{\hat{\bm{n}}} \bm{W}_f \bm{T}_f \bm{M}_m^{-1}\bm{T}_m\bm{W}_m\\
&=  \bm{W}_f \bm{T}_f \bm{M}_m^{-1}\bm{T}^T_m\bm{W}_m\diag{\hat{\bm{n}}_i} = \bm{P}_f^T\bm{T}^T_m \tilde{\bm{B}}_i = \LRp{\tilde{\bm{B}}_i{\bm{E}}_m}^T
\end{align*}
where we have used that $\bm{W}_m$ is diagonal and that the outward normal $\hat{\bm{n}}_i$ on the reference element is constant over each face.  This implies skew-symmetry of the decoupled SBP operator and entropy stability of the overall system.

@@

While the formulation presented above is convenient for analysis, it is computationally expensive to implement.  However, using properties of the mortar matrices $\bm{E}_m, \tilde{\bm{E}}_m$, we can rewrite the above formulation in a way which reflects a traditional mortar-based finite element implementation.  In other words, we wish to implement (\ref{eq:mortar}) such that the only modification from the implementations in \cite{chan2017discretely} is a pre-processing step on mortar faces.

We first note that, since $\bm{E}_m$ maps from surface nodes to mortar nodes, $\bm{V}_m = \bm{E}_m \bm{V}_f$.  Thus, we can rewrite the numerical flux contribution as
\begin{align*}
 \bm{V}_m^T\tilde{\bm{B}}_i\bm{f}_i^* &= \bm{V}_f^T \bm{E}_m^T \bm{W}_m\diag{\hat{\bm{n}}_i} \bm{f}_i^* = \bm{V}_f^T  \bm{P}_f^T\bm{T}_m^T\bm{W}_m\diag{\hat{\bm{n}}_i} \bm{f}_i^*\\
 &= \bm{V}_f^T \bm{W}_f \diag{\hat{\bm{n}}_i} \bm{T}_m\bm{M}_m^{-1}\bm{T}_m^T\bm{W}_m \bm{f}_i^* =  \bm{V}_f^T \bm{B}_i \tilde{\bm{E}}_m\bm{f}_i^*.
\end{align*}

We can similarly reformulate the remaining mortar contributions within (\ref{eq:mortar}).  We decompose $\bm{F}_S$ into interactions between volume nodes, surface nodes, and mortar nodes
\[
\bm{F}_S = \begin{bmatrix}
\bm{F}_S^{vv} & \bm{F}_S^{vs} & \bm{F}_S^{vm}\\
\bm{F}_S^{sv} & \bm{F}_S^{ss} & \bm{F}_S^{sm}\\
\bm{F}_S^{mv} & \bm{F}_S^{ms} & \bm{F}_S^{mm}
\end{bmatrix}
\]
Then, the on-element contributions to (\ref{eq:mortar}) can be expanded as follows
\begin{align*}
&\begin{bmatrix} \bm{V}_q \\ \bm{V}_f \\ \bm{V}_m \end{bmatrix}^T
\LRp{\begin{bmatrix}
\bm{Q}_i-\bm{Q}_i^T & \bm{E}^T\bm{B}_i &\\
-\bm{B}_i\bm{E} &  & \bm{B}_i\tilde{\bm{E}}_m \\
& -\tilde{\bm{B}}_i{\bm{E}}_m & 
\end{bmatrix} \circ \bm{F}_S}\bm{1} \\
&= 
\begin{bmatrix} \bm{V}_q \\ \bm{V}_f \end{bmatrix}^T
\LRp{\begin{bmatrix}
\bm{Q}_i-\bm{Q}_i^T & \bm{E}^T\bm{B}_i\\
-\bm{B}_i\bm{E} & \\
\end{bmatrix} \circ \bm{F}_S}\bm{1} \\
&+ \bm{V}_f^T \LRp{\bm{B}_i\tilde{\bm{E}}_m\circ \bm{F}_S^{sm}}\bm{1} - \bm{V}_m^T \LRp{ \tilde{\bm{B}}_i\bm{E}_m \circ \bm{F}_S^{ms}}\bm{1}
\end{align*}
Since multiplication by diagonal matrices $\bm{B}_i, \tilde{\bm{B}}_i$ is associative under the Hadamard product, the latter terms can be rewritten as 
\begin{align*}
 \bm{V}_f^T \LRp{\bm{B}_i\tilde{\bm{E}}_m\circ \bm{F}_S^{sm}}\bm{1} &=  \bm{V}_f^T \bm{B}_i \LRp{\tilde{\bm{E}}_m\circ \bm{F}_S^{sm}}\bm{1}\\
\bm{V}_m^T \LRp{ \tilde{\bm{B}}_i\bm{E}_m \circ \bm{F}_S^{ms}}\bm{1} &= \bm{V}_m^T \tilde{\bm{B}}_i \LRp{ \bm{E}_m \circ \bm{F}_S^{ms}}\bm{1} = \bm{V}_f^T \bm{B}_i \tilde{\bm{E}}_m \LRp{ \bm{E}_m \circ \bm{F}_S^{ms}}\bm{1} 
\end{align*}
Then, (\ref{eq:mortar}) can be rewritten as 
\begin{align}
\bm{M}\td{\bm{u}_N}{t} &+ 
\begin{bmatrix} \bm{V}_q \\ \bm{V}_f \end{bmatrix}^T
\LRp{\begin{bmatrix}
\bm{Q}_i-\bm{Q}_i^T & \bm{E}^T\bm{B}_i\\
-\bm{B}_i\bm{E} & \\
\end{bmatrix} \circ \bm{F}_S}\bm{1} + \bm{V}_f^T\bm{B}_i \tilde{\bm{f}}^*_i = 0 \label{eq:mortar2}\\
\tilde{\bm{f}}^*_i &= \tilde{\bm{E}}_m\bm{f}^*_i + \LRp{\tilde{\bm{E}}_m\circ \bm{F}_S^{sm}}\bm{1} - \tilde{\bm{E}}_m \LRp{ \bm{E}_m \circ \bm{F}_S^{ms}}\bm{1} 
\nonumber
\end{align}
If the surface and mortar nodes are identical, then $\bm{E}_m = \tilde{\bm{E}}_m = \bm{I}$ and $\LRp{\tilde{\bm{E}}_m \circ \bm{F}_S^{sm}}\bm{1} = \LRp{ \bm{E}_m \circ \bm{F}_S^{ms}}\bm{1}$.  The correction terms within $\tilde{\bm{f}}^*$ cancel out, and we recover the original scheme from \cite{chan2017discretely}.

Note that in (\ref{eq:mortar2}), the only interactions between surface and mortar nodes occurs within the computation of $\tilde{\bm{f}}^*_i$, which can be done solely on mortar interfaces.  The computational structure of 
\begin{enumerate}
\item Compute volume contributions to the RHS
\item Communicate surface contributions to mortar interfaces and compute $\tilde{\bm{f}}^*_i$.
\item Communicate $\tilde{\bm{f}}^*_i$ back to each element.  
\end{enumerate}
%\bm{M}\td{\bm{u}_N}{t} + \begin{bmatrix} \bm{V}_q \\ \bm{V}_f \\ \bm{V}_m \end{bmatrix}^T
%\LRp{\begin{bmatrix}
%\bm{Q}_i-\bm{Q}_i^T & \bm{E}^T\bm{B}_i &\\
%-\bm{B}_i\bm{E} &  & \bm{B}_i\tilde{\bm{E}}_m \\
%& -\tilde{\bm{B}}_i{\bm{E}}_m & 
%\end{bmatrix} \circ \bm{F}_S}\bm{1} + \bm{V}_m^T\tilde{\bm{B}}_i\bm{f}_i^* = 0.

@@

\section{Error estimates}

\note{Weak derivative vs strong derivative error estimates.  SBP yields that the skew-symmetric form is equivalent to weak form.}

\note{Line DG approach can work, but it's flipped: we increase quadrature strength in the direction \emph{orthogonal} to derivative instead.  May improve accuracy over GLL by giving equivalence w/skew form.}

\section{Numerical experiments}

\subsection{Gauss Lobatto quadrature}

\subsection{Gauss quadrature}
